---
title: Second Tutorial for R4DS
author: Sruthi Gandhi and David Kane
tutorial:
  id: r4ds-2
output:
  learnr::tutorial:
    df_print: default
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: 'This tutorial for "R for Data Science" covers Chapter 9: Layers, Chapter 10: Exploratory Data Analysis, Chapter 11: Communication, and Chapter 20: Spreadsheets. Exercises practice ggplot2 layers, faceting, themes, and reading Excel files with readxl.'
---

```{r setup, include = FALSE}
library(learnr)
library(tutorial.helpers)

library(tidyverse)
library(readxl)
library(dbplyr)

knitr::opts_chunk$set(echo = FALSE)
options(tutorial.exercise.timelimit = 600,
        tutorial.storage = "local")


cheese_tibble <- read_excel("data/cheeses.xlsx") |>
  filter(!is.na(country)) |>
  add_count(country, name = "country_total") |>
  filter(country_total >= sort(unique(country_total), decreasing = TRUE)[3]) |>
  filter(!is.na(milk)) |>
  count(country, milk) |>
  group_by(country) |>
  arrange(country, desc(n))

wine_tibble <- read_excel("data/wine.xlsx")|>
  select(Phenols, Color, Flavanoids)
```

```{r info-section, child = system.file("child_documents/info_section.Rmd", package = "tutorial.helpers")}
```

## Introduction
###

This section covers key concepts from [Chapter 9: Layers](https://r4ds.hadley.nz/layers.html), [Chapter 10: Exploratory Data Analysis](https://r4ds.hadley.nz/eda.html), [Chapter 11: Communication](https://r4ds.hadley.nz/communication.html), and [Chapter 20: Spreadsheets](https://r4ds.hadley.nz/spreadsheets.html) from [*R for Data Science (2e)*](https://r4ds.hadley.nz/) by Hadley Wickham, Mine Çetinkaya-Rundel, and Garrett Grolemund.

You will learn about working with categorical and continuous variables using different types of variation and exploratory data analysis using core packages such as **[readxl](https://readxl.tidyverse.org/)**, **[dbplyr](https://dbplyr.tidyverse.org/)**, and **[ggplot2](https://ggplot2.tidyverse.org/)**. Key **ggplot2** techniques include faceting plots into panels, flipping axes, and adding jitter to avoid overplotting.

We recommend using an in-environment AI agent such as [Gemini CLI](https://github.com/google-gemini/gemini-cli) or [Claude Code](https://claude.ai/code), which can read and edit your files directly. Instructions are written with these tools in mind. You may also use a chat-based AI, but you will need to copy/paste code and data context manually.


### Exercise 1

Create a new repo using `codespace-starter` as a template and name it `r4ds-2`.

If you haven't already, open the `r4ds-2` repo on GitHub and create a Codespace from it.

You need two R Terminals: one for running this tutorial and one for your exercises. Open a second R Terminal using the dropdown button.

Create a new file and save it as `analysis.qmd`. Add a YAML header at the top with a title (`"Analyzing Cheese and Wine"`) and your name as author. In a bash terminal, run:

```
quarto render analysis.qmd
```

Open `analysis.html` in a browser by right-clicking it in the File Explorer and selecting **Open with Live Server**. Keep this tab open — it will refresh automatically whenever you re-render.

Create a `.gitignore` file with `analysis_files` on the first line and then a blank line. Save and push.

In the R Terminal, run:

```
show_file(".gitignore")
```

If that fails, it is probably because you have not yet loaded `library(tutorial.helpers)` in the R Terminal.

CP/CR.

```{r introduction-1}
question_text(NULL,
    answer(NULL, correct = TRUE),
    allow_retry = TRUE,
    try_again_button = "Edit Answer",
    incorrect = NULL,
    rows = 3)
```

###

**googlesheets4** uses v4 of the [Google Sheets API v4](https://developers.google.com/sheets/api/) to provide an R interface to Google Sheets, hence the name.
You can create a brand new sheet with [`gs4_create()`](https://googlesheets4.tidyverse.org/reference/gs4_create.html) or write to an existing sheet with [`sheet_write()`](https://googlesheets4.tidyverse.org/articles/write-sheets.html) and friends.

### Exercise 2

In your QMD, put `library(tidyverse)`, `library(readxl)`, and `library(dbplyr)` in a new code chunk. In a bash terminal, run `quarto render analysis.qmd`. Check your Live Server tab — it should refresh automatically.

Notice that the file does not look good because the code is visible and there are annoying messages. To take care of this, add `#| message: false` to remove all the messages in this code chunk. Also add the following to the YAML header to remove all code echoes from the HTML:

```
execute:
  echo: false
```

In the R Terminal, run:

```
show_file("analysis.qmd", chunk = "Last")
```

CP/CR.

```{r introduction-2}
question_text(NULL,
    answer(NULL, correct = TRUE),
    allow_retry = TRUE,
    try_again_button = "Edit Answer",
    incorrect = NULL,
    rows = 6)
```

###

You can read a single worksheet from a spreadsheet with the `sheet` argument in `read_excel()` from the **readxl** package.

### Exercise 3

In a bash terminal, run `quarto render analysis.qmd`. Check your Live Server tab.

Note that this causes `library(tidyverse)`, `library(readxl)`, and `library(dbplyr)` to be copied down to the R Terminal and then executed.

CP/CR.

```{r introduction-3}
question_text(NULL,
    answer(NULL, correct = TRUE),
    allow_retry = TRUE,
    try_again_button = "Edit Answer",
    incorrect = NULL,
    rows = 3)
```

###

You can use `excel_sheets()` to get information on all worksheets in an Excel spreadsheet, and then read the one(s) you're interested in.

### Exercise 4

From the R Terminal, run these three commands:

`getwd()`
`dir.create("data")`
`list.files()`

This will create a `data` directory in your project. This is a good place to store any data that you are working with.

CP/CR.

```{r introduction-4}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

Your answer should look something like this, although your path will be different.

````
> getwd()
[1] "/workspaces/r4ds-2"
> dir.create("data")
> list.files()
 [1] "analysis_files" "analysis.html"  "analysis.qmd"   "data"
>
````

## Cheese
###

The [cheese.csv](https://github.com/rfordatascience/tidytuesday/blob/main/data/2024/2024-06-04/readme.md) dataset is a [TidyTuesday](https://github.com/rfordatascience/tidytuesday) dataset on Github which contains data scraped from [cheese.com](https://www.cheese.com) about data from 2046 different types of cheese. Our spreadsheet `data/cheeses.xlsx` contains this data in a spreadsheet format.

### Exercise 1

We begin by downloading our spreadsheet `cheeses.xlsx` directly from GitHub to the `data` directory using `download.file()`.

In the R Terminal, run:

```
download.file(
  "https://github.com/PPBDS/vscode.tutorials/raw/refs/heads/main/inst/tutorials/r4ds-2/data/cheeses.xlsx",
  destfile = "data/cheeses.xlsx",
  mode = "wb"
)
```

CP/CR.

```{r cheese-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 8)
```

###

`download.file()` saves a file from a URL to a local path. The `mode = "wb"` argument is required for binary files like `.xlsx` — without it, text-mode line-ending conversion can corrupt the file.

### Exercise 2

Ask AI how to read a `.xlsx` spreadsheet called `data/cheeses.xlsx` using `read_excel()`. Add this code to a new code chunk in your QMD. In a bash terminal, run `quarto render analysis.qmd`. Check your Live Server tab.

CP/CR.

```{r cheese-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

Our result:

```{r cheese-2-test}
read_excel("data/cheeses.xlsx")
```

`read_excel()` can read files with either `.xls` and `.xlsx` format. It guesses the file type based on the input.

### Exercise 3

Use the following prompt with your AI to filter and summarize the cheese data. Add the resulting code as a replacement of your current code in the same code chunk in your QMD. In a bash terminal, run `quarto render analysis.qmd`.

In the R Terminal, run:

```
show_file("analysis.qmd", chunk = "Last")
```

CP/CR.

Here's the prompt we used:

Write R code that starts with `read_excel("data/cheeses.xlsx")` and pipes it through several operations. First, filter out rows where country is NA using `filter(!is.na(country))`, then add a column showing the total count for each country using `add_count(country, name = "country_total")`. Next, keep only the top 3 countries by filtering where the `country_total` is greater than or equal to the 3rd highest value using `filter(country_total >= sort(unique(country_total), decreasing = TRUE)[3])`. After that, filter out any NA values in the `milk` column with `!is.na(milk)`, then count the combinations of `country` and `milk` type using `count(country, milk)`, group the results by `country`, and finally arrange them by country name and then by count in descending order.

```{r cheese-3}
question_text(NULL,
    answer(NULL, correct = TRUE),
    allow_retry = TRUE,
    try_again_button = "Edit Answer",
    incorrect = NULL,
    rows = 6)
```

###

Our result:

```{r cheese-3-test}
read_excel("data/cheeses.xlsx") |>
  filter(!is.na(country)) |>
  add_count(country, name = "country_total") |>
  filter(country_total >= sort(unique(country_total), decreasing = TRUE)[3]) |>
  filter(!is.na(milk)) |>
  count(country, milk) |>
  group_by(country) |>
  arrange(country, desc(n))
```

If your result doesn't match ours, replace your chunk code with the following before moving on:

<pre><code>```{r}
read_excel("data/cheeses.xlsx") |>
  filter(!is.na(country)) |>
  add_count(country, name = "country_total") |>
  filter(country_total >= sort(unique(country_total), decreasing = TRUE)[3]) |>
  filter(!is.na(milk)) |>
  count(country, milk) |>
  group_by(country) |>
  arrange(country, desc(n))
```</code></pre>

`add_count()` adds a column of group counts without collapsing the data — equivalent to `group_by()` followed by `add_tally()`. Use it when you need the count as a new column for downstream filtering.

### Exercise 4

Within the recent code chunk, add the option `#| cache: true` and assign the result to `cheese_tibble`. In a bash terminal, run `quarto render analysis.qmd`. This first render takes noticeably longer — Quarto runs the code and writes the cache to disk; subsequent renders skip that code and load from cache instead.

In the R Terminal, run:

```
show_file("analysis.qmd", chunk = "Last")
```

CP/CR.

```{r cheese-4}
question_text(NULL,
    answer(NULL, correct = TRUE),
    allow_retry = TRUE,
    try_again_button = "Edit Answer",
    incorrect = NULL,
    rows = 6)
```

###

Quarto's `#| cache: true` option saves a chunk's result to disk. Add `analysis_cache` to your `.gitignore` so the cache files don't end up on GitHub. `count()` outputs a tibble with variable values and their observation counts; the resulting `n` column can be mapped directly to aesthetics like `size`, `y`, or `fill` in **ggplot2**.

### Exercise 5

Ask AI to pipe `cheese_tibble` into a plot that shows the number of cheeses by milk type, faceted by country, with flipped axes. Add this code to a new code chunk in your QMD. In a bash terminal, run `quarto render analysis.qmd`. Check your Live Server tab.

In the R Terminal, run:

```
show_file("analysis.qmd", chunk = "Last")
```

CP/CR.

```{r cheese-5}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

Our plot:

```{r cheese-5-test}
cheese_tibble |>
  ggplot(aes(x = milk, y = n, fill = milk)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ country, scales = "fixed") +
  coord_flip() +
  scale_y_sqrt() +
  theme_gray() +
  labs(
    title = "Milk Types Used in Cheeses by Country",
    subtitle = "Italy has the most variety of cheese",
    caption = "Source: cheese.com",
    x = "Type of Milk",
    y = "Number of Cheeses (sqrt scale)"
  )
```

**Faceting** splits a plot into panels, one per level of a variable. `coord_flip()` swaps the x and y axes, making long category labels easier to read.

## Wine
###

The `wine` dataset contains the results of a chemical analysis of wines grown in a specific area of Italy. Three types of wine are represented in the 178 samples, with the results of 13 chemical analyses recorded for each sample. This data is from the UCI Machine Learning Repository. Our spreadsheet `data/wine.xlsx` contains this data in a spreadsheet format.

### Exercise 1

We begin by downloading our spreadsheet `wine.xlsx` directly from GitHub to the `data` directory using `download.file()`.

In the R Terminal, run:

```
download.file(
  "https://github.com/PPBDS/vscode.tutorials/raw/refs/heads/main/inst/tutorials/r4ds-2/data/wine.xlsx",
  destfile = "data/wine.xlsx",
  mode = "wb"
)
```

CP/CR.

```{r wine-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

You can write data back to disk as an Excel file using the `write_xlsx()` function from the **writexl** package.

### Exercise 2

Ask AI how to read a `.xlsx` spreadsheet called `data/wine.xlsx` using `read_excel()`. Add this code to a new code chunk in your QMD. In a bash terminal, run `quarto render analysis.qmd`. Check your Live Server tab.

CP/CR.

```{r wine-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

Our result:

```{r wine-2-test}
read_excel("data/wine.xlsx")
```

`scale_y_continuous()` modifies the y-axis for continuous data — use it to set breaks, labels, limits, or apply a transformation like `trans = "log"`.

### Exercise 3

Ask AI how to pipe `read_excel("data/wine.xlsx")` into `select()` to keep only the `Phenols`, `Color`, and `Flavanoids` columns. Add this code as a replacement of your current code in the same code chunk in your QMD. In a bash terminal, run `quarto render analysis.qmd`.

In the R Terminal, run:

```
show_file("analysis.qmd", chunk = "Last")
```

CP/CR.

```{r wine-3}
question_text(NULL,
    answer(NULL, correct = TRUE),
    allow_retry = TRUE,
    try_again_button = "Edit Answer",
    incorrect = NULL,
    rows = 6)
```

###

Our result:

```{r wine-3-test}
read_excel("data/wine.xlsx") |>
  select(Phenols, Color, Flavanoids)
```

If your result doesn't match ours, replace your chunk code with the following before moving on:

<pre><code>```{r}
read_excel("data/wine.xlsx") |>
  select(Phenols, Color, Flavanoids)
```</code></pre>

`select()` from **dplyr** keeps only the named columns. You can also drop columns with a minus sign, e.g., `select(-unwanted_col)`.

### Exercise 4

Within the recent code chunk, add the option `#| cache: true` and assign the result to `wine_tibble`. In a bash terminal, run `quarto render analysis.qmd`.

In the R Terminal, run:

```
show_file("analysis.qmd", chunk = "Last")
```

CP/CR.

```{r wine-4}
question_text(NULL,
    answer(NULL, correct = TRUE),
    allow_retry = TRUE,
    try_again_button = "Edit Answer",
    incorrect = NULL,
    rows = 6)
```

###

Covariation is the tendency for the values of two or more variables to vary together in a related way. The best way to spot covariation is to visualize the relationship between variables using a scatter plot.

### Exercise 5

Ask AI to create a scatter plot from `wine_tibble` with `Phenols` on the x-axis and `Flavanoids` on the y-axis, with color mapped to `Color`. Include jittered points, a LOESS smooth curve, and label the maximum `Flavanoids` value as "Outlier" in magenta. Use a blue-to-orange color gradient and a classic theme. Add this code to a new code chunk in your QMD. In a bash terminal, run `quarto render analysis.qmd`. Check your Live Server tab.

In the R Terminal, run:

```
show_file("analysis.qmd", chunk = "Last")
```

CP/CR.

```{r wine-5}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 12)
```

###

Our plot:

```{r wine-5-test}
wine_tibble |>
  ggplot(aes(x = Phenols, y = Flavanoids, color = Color)) +
  geom_point(size = 2, alpha = 0.5) +
  geom_jitter() +
  geom_smooth(method = "loess", se = TRUE, color = "black") +
  geom_text(
    data = \(d) d |> filter(Flavanoids == max(Flavanoids)),
    aes(label = "Outlier"),
    nudge_y = 0.2,
    color = "magenta",
    fontface = "bold"
  ) +
  scale_color_gradient(low = "blue", high = "orange") +
  theme_classic() +
  labs(
    title = "Wine: Flavanoids vs Phenols",
    x = "Phenols",
    y = "Flavanoids"
  )
```

`geom_jitter()` is shorthand for `geom_point(position = "jitter")` — it adds a small amount of random noise to each point to avoid overplotting in dense regions.

## Summary
###

This section covered key concepts from [Chapter 9: Layers](https://r4ds.hadley.nz/layers.html), [Chapter 10: Exploratory Data Analysis](https://r4ds.hadley.nz/eda.html), [Chapter 11: Communication](https://r4ds.hadley.nz/communication.html), and [Chapter 20: Spreadsheets](https://r4ds.hadley.nz/spreadsheets.html) from [*R for Data Science (2e)*](https://r4ds.hadley.nz/) by Hadley Wickham, Mine Çetinkaya-Rundel, and Garrett Grolemund.

You learned about working with categorical and continuous variables using different types of variation and exploratory data analysis with core packages such as **[readxl](https://readxl.tidyverse.org/)**, **[dbplyr](https://dbplyr.tidyverse.org/)**, and **[ggplot2](https://ggplot2.tidyverse.org/)**. Key **ggplot2** techniques included faceting plots into panels, flipping axes, and adding jitter to avoid overplotting.

### Exercise 1

In a bash terminal, run `quarto render analysis.qmd` to ensure that everything works. Check your Live Server tab — the resulting HTML page should be attractive, showing clean versions of your plots.

In the R Terminal, run:

```
show_file("analysis.qmd")
```

CP/CR.

```{r summary-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 30)
```

###

Spreadsheet data in `.xlsx` format can be read into R using `read_excel()` from the **readxl** package.

### Exercise 2

Publish your rendered QMD to GitHub Pages. In a bash terminal, run:

````
quarto publish gh-pages analysis.qmd
````

If the bash terminal is still running from rendering, stop it first using `Ctrl/Cmd + C`.

Copy/paste the resulting URL below.

```{r summary-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 1)
```

###

Publishing the QMD creates a public URL anyone can use to view the rendered project.

### Exercise 3

Commit and push all your files. Copy/paste the URL to your GitHub repo.

```{r summary-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

GitHub repositories are where people can view the code you used to create your project.

```{r download-answers, child = system.file("child_documents/download_answers.Rmd", package = "tutorial.helpers")}
```
