---
title: Analyzing US Census Data 
author: David Kane, Ansh Pathak, and Paramanyu Gupta
tutorial:
  id: tidycensus-1
output: 
  learnr::tutorial:
    df_print: default
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: "Overview tutorial for the 'Analyzing US Census Data' book. Walks through three case studies on Texas income, California bachelors degrees, and California median age, using tidycensus to pull ACS data and ggplot2 with geom_sf() for choropleth maps."
---

```{r setup, include = FALSE}
library(learnr)
library(tutorial.helpers)
library(tidycensus)
library(tidyverse)
library(knitr)


# income_tx <- get_acs(
#   geography = "county",
#   variables = "B19013_001",
#   state = "TX",
#   year = 2020,
#   geometry = TRUE
# )
# write_rds(income_tx, "data/income_tx.rds")

income_tx <- read_rds("data/income_tx.rds")


# edu_ca <- get_acs(
#   geography = "county",
#   variables = c("B15003_001", "B15003_022", "B15003_023", "B15003_024", "B15003_025"),
#   state = "CA",
#   year = 2020,
#   geometry = TRUE,
#   summary_var = "B15003_001"
# )
#  write_rds(edu_ca, "data/edu_ca.rds")

edu_ca <- read_rds("data/edu_ca.rds")


# age_ca <- get_acs(
#   geography = "county",
#   variables = c(median_age = "B01002_001", population = "B01003_001"),
#   state = "CA",
#   year = 2020,
#   geometry = FALSE
# )
# write_rds(age_ca, "data/age_ca.rds")

age_ca <- read_rds("data/age_ca.rds")

knitr::opts_chunk$set(echo = FALSE)
options(tutorial.exercise.timelimit = 600, 
        tutorial.storage = "local") 	
```



```{r info-section, child = system.file("child_documents/info_section.Rmd", package = "tutorial.helpers")}
```


## Introduction

<!-- DK: Overview introduction. Link to the book.  -->


This tutorial covers an overview of [Analyzing US Census Data](https://walker-data.com/census-r/index.html) by Kyle Walker. You will learn about using the [**tidycensus**] package for collecting, interacting, and plotting US Census data. You will mainly focus on collecting data from the Decennial Census and the American Community Survey (ACS).

### Exercise 1

Create a new repo using `codespace-starter` as a template and name it `tidycensus-plots`.

If you haven't already, open the `tidycensus-plots` repo on GitHub and create a Codespace from it.

You need two R Terminals: one for running this tutorial and one for your exercises.

Create a new Quarto Document and save it as `TexasIncome.qmd`. Add a YAML header at the top with a title (`"TexasIncome"`) and your name as author. Render the document using `Cmd/Ctrl + Shift + K`.

Create a `.gitignore` file with `*_files` on the first line and then a blank line. Save and push.

In the R Terminal, run:

```         
show_file(".gitignore")
```

If that fails, it is probably because you have not yet loaded `library(tutorial.helpers)` in the R Terminal.

CP/CR.

```{r introduction-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

### 

Professionals keep their data science work in the cloud because laptops fail.

### Exercise 2

In your QMD, put `library(tidyverse)`, `library(tidycensus)`, and `library(sf)` in a new code chunk. Press `Ctrl/Cmd + Shift + K` to render the file

Notice that the file does not look good because the code is visible and there are annoying messages. To take care of this, add `#| message: false` to remove all the messages in this `setup` chunk. Also add the following to the YAML header to remove all code echoes from the HTML:

```         
execute: 
  echo: false
```

In the R Terminal, run:

```         
show_file("TexasIncome.qmd", chunk = "Last")
```

CP/CR.

```{r introduction-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```

### 

Render again. Everything looks nice, albeit empty, because we have added code to make the file look better and more professional.

### Exercise 3

Place your cursor in the QMD file on the `library(tidyverse)` line. Use `Cmd/Ctrl + Enter` to execute that line.

Note that this causes the entire code cell to be copied down to the R Terminal and then executed. 

CP/CR.

```{r introduction-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

This is how professionals work, they type code in the quarto document and send it down to the console to execute it!

## Texas Income

<!-- DK: Clean up as we discussed. No wasted words. -->

A critical part of the Census data analysis process is data visualization, where an analyst examines patterns and trends found in their data graphically. This first section illustrates some examples for getting started with exploratory Census data visualization with [**ggplot2**](https://ggplot2.tidyverse.org/). You will be using the `get_acs()` and the `geom_sf()` functions as well.

### Exercise 1

Ask AI to use **tidycensus** to get data on the median household income for all counties in Texas for 2020.

Put what it gives you in a new code chunk and do `Cmd/Ctrl + Enter` to send it to the R Terminal.

CP/CR.

```{r texas-income-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```

###

Our answer:

````
income_tx <- get_acs(
  geography = "county",
  variables = "B19013_001",
  state = "TX",
  year = 2020,
  geometry = FALSE
)
````

###

<!-- DK: Knowledge drop. -->

The `geometry` argument in `get_acs()` controls whether spatial boundary data is downloaded. Setting `geometry = TRUE` downloads county/tract shapes as polygons, which allows you to create maps with `geom_sf()`.

### Exercise 2

Copy and paste our code to replace it with what you have.


In the R Terminal, run:

```         
show_file("TexasIncome.qmd", chunk = "Last")
```

CP/CR.

```{r texas-income-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```

###

`get_acs()` is part of the **tidycensus** package and allows downloading American Community Survey (ACS) data.

### Exercise 3

Edit your QMD, to change the value of `geometry` to `TRUE`.

Take your code chunk and do `Cmd/Ctrl + Enter` to send it to the R Terminal.

In the R Terminal, run:

```         
show_file("TexasIncome.qmd", chunk = "Last")
```

CP/CR.

```{r texas-income-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```

###

````
> income_tx <- get_acs(
+   geography = "county",
+   variables = "B19013_001",
+   state = "TX",
+   year = 2020,
+   geometry = TRUE
+ )
Getting data from the 2016-2020 5-year ACS
Downloading feature geometry from the Census website.  To cache shapefiles for use in future sessions, set `options(tigris_use_cache = TRUE)`.
Using FIPS code '48' for state 'TX'
  |========================================================================================================================| 100%
>
````

###

The `geometry = TRUE` argument returns spatial polygons, useful for maps and spatial analysis.

### Exercise 4

Run `glimpse()` on `income_tx` in the R Terminal.

CP/CR.

```{r texas-income-4}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

<!-- DK: Knowledge drop? -->

The `glimpse()` function from **dplyr** shows the structure of your data frame transposed, making it easy to scan all column names and types at once. For Census data, this helps you quickly verify you got the right variables and see what additional columns like `GEOID` and `NAME` were automatically included.

### Exercise 5

Run `summary(income_tx)` in the console.

CP/CR.


```{r texas-income-5}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

<!-- DK: Chance to define the estimate and moe. -->

ACS data includes both an `estimate` column (the survey's best guess) and a `moe` column (margin of error). The margin of error tells you the range of uncertainty. Smaller margins mean more reliable estimates, which typically occurs in areas with larger populations.

### Exercise 6

Now, type `income_tx` in the console.

CP/CR the first few lines.

```{r texas-income-6}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

`ggplot2` is ideal for Census data due to its support for basic charts, group comparisons, and geospatial visualizations.


### Exercise 7

Now copy the first few lines into your AI and say that you are working with **tidyverse**. Tell it to take the data in `income_tx` and make a choropleth map of median household income.


Put what it gives you in a new code chunk and do `Cmd/Ctrl + Enter` to send it to the console

CP/CR.

```{r texas-income-7}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```

###

Our answer:

````
ggplot(income_tx) +
  geom_sf(aes(fill = estimate), color = "white", size = 0.2)
````

### Exercise 8

Now, tell the AI to color counties by the estimate column and add an approprite title and theme. 


Replace your code with this code in your chunk and do `Cmd/Ctrl + Enter` to send it to the console

CP/CR.

```{r texas-income-8}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```
###


When mapping spatial Census data, use `geom_sf()` in `ggplot2` and apply `fill = estimate` to show choropleth patterns.


### Exercise 9

Here is our code. It is okay if your code is different. That will happen when using AI!
Replace your code with what it gave you using this code: 

````
ggplot(income_tx) +
  geom_sf(aes(fill = estimate), color = "white", size = 0.2) +
  scale_fill_viridis_c(option = "plasma", name = "Median Income") +
  labs(
    title = "Median Household Income by County, Texas (2020)",
    caption = "Data source: ACS 5-year estimates"
  ) +
  theme_minimal()
````

###

`geom_sf()` plots spatial data stored as simple features (`sf` objects). The `fill` aesthetic maps color to median income. The `viridis` color scale provides perceptually uniform colors for better interpretation.
```{r}
#| message: false
ggplot(income_tx) +
  geom_sf(aes(fill = estimate), color = "white", size = 0.2) +
  scale_fill_viridis_c(option = "plasma", name = "Median Income") +
  labs(
    title = "Median Household Income by County, Texas (2020)",
    caption = "Data source: ACS 5-year estimates"
  ) +
  theme_minimal()
```

### Exercise 10


In the R Terminal, run the following command to display the last chunk of your `.qmd` file: CP/CR

```         
show_file("TexasIncome.qmd", chunk = "Last")
```

```{r texas-income-10}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 15)
```

###

The `show_file()` function from **tutorial.helpers** is a convenient way to check the contents of files without leaving R. It helps confirm that your edits were saved properly.


## California Bachelors Degree

In this activity, we will be using **ggplot**, `geom_sf`, and be modifying data to turn it into percentages.


### 

Create a new Quarto Document and save it as `CaliforniaBachelors.qmd`. Add a YAML header at the top with a title (`"CaliforniaBachelors"`) and your name as author. Render the document using `Cmd/Ctrl + Shift + K`.

In this exercise, we’ll get the percentage of adults with a bachelor’s degree or higher in each California county.

### Exercise 1

In your QMD, put `library(tidyverse)`, `library(tidycensus)`, and `library(sf)` in a new code chunk. Press Ctrl/Cmd + Shift + K to render the file

Notice that the file does not look good because the code is visible and there are annoying messages. To take care of this, add `#| message: false` to remove all the messages in this `setup` chunk. Also add the following to the YAML header to remove all code echoes from the HTML:

```
execute:
  echo: false
```

In the R Terminal, run:

```
show_file("CaliforniaBachelors.qmd", chunk = "Last")
```

CP/CR.

```{r california-bachelors-degree-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```

### 

Render again. Everything looks nice, albeit empty, because we have added code to make the file look better and more professional.

### Exercise 2

Place your cursor in the QMD file on the `library(tidyverse)` line. Use `Cmd/Ctrl + Enter` to execute that line.

Note that this causes `library(tidyverse)` to be copied down to the R Terminal and then executed. 

CP/CR.

```{r california-bachelors-degree-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

Educational attainment variables in the Census are nested hierarchically—table B15003 contains 25 separate variables from "No schooling" through "Doctorate degree." To calculate percentages for bachelor's degree or higher, you need to identify and sum the specific variable codes (B15003_022 through B15003_025) that represent those education levels.

### Exercise 3

Ask AI to generate R code that uses tidycensus to get educational attainment variables for all California counties in 2020 and save it in a variable called `edu_ca`

Put what it gives you in a new code chunk and do `Cmd/Ctrl + Enter` to send it to the console

CP/CR


```{r california-bachelors-degree-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```


###

Our code:

````
edu_ca <- get_acs(
  geography = "county",
  variables = c("B15003_001", "B15003_022", "B15003_023", "B15003_024", "B15003_025"),
  state = "CA",
  year = 2020,
  geometry = FALSE,
  summary_var = "B15003_001"
)
````

### Exercise 4

Using our code, set `geometry = TRUE`.

Take your code chunk and do `Cmd/Ctrl + Enter` to send it to the R Terminal.

In the R Terminal, run:

```         
show_file("CaliforniaBachelors.qmd", chunk = "Last")
```

CP/CR.

```{r california-bachelors-degree-4}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```

###

````
edu_ca <- get_acs(
  geography = "county",
  variables = c("B15003_001", "B15003_022", "B15003_023", "B15003_024", "B15003_025"),
  state = "CA",
  year = 2020,
  geometry = TRUE,
  summary_var = "B15003_001"
)
````

The American Community Survey (ACS) provides annual demographic, economic, and housing data based on samples, while the Decennial Census gives a complete count every 10 years.

### Exercise 5

Let's explore the data, run `glimpse(edu_ca)` in the console.

CP/CR.

```{r california-bachelors-degree-5}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

When requesting multiple variables from the same table, **tidycensus** returns data in long format with one row per geography-variable combination. The `summary_var` argument adds a column with totals, useful for calculating percentages later.

### Exercise 6

Let's get some quick stats for each column of the data. 

Run `summary(edu_ca)` in the console.

CP/CR.


```{r california-bachelors-degree-6}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

Census education tables report completed degrees, not enrollment. Someone with a master's degree appears in both the bachelor's and master's variables, so you must be careful not to double-count when summing across education levels.

### Exercise 7

We’ll now make a choropleth map of bachelor’s degree attainment across California counties.

Send our code to the console from the previous exercise.

Now, type `edu_ca` in the console.

CP/CR the first few lines.

```{r california-bachelors-degree-7}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

Geometry must be `TRUE` if we want to map later.

### Exercise 8

Now copy the first few lines into your AI and say that you are working with tidyverse. Tell it to take the data in `edu_ca` and make a choropleth map of at least bachelor’s degree attainment or higher across California counties and use `mutate()` to calculate the percentage of the population with at least a bachelor’s degree and pipe it into a `ggplot()`

Put what it gives you in a new code chunk and do `Cmd/Ctrl + Enter` to send it to the console

CP/CR

```{r california-bachelors-degree-8}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```

###

Our Code:

````
edu_ca <- edu_ca |>
  group_by(GEOID) |>
  summarize(
    percent = 100 * sum(estimate[variable != "B15003_001"]) / unique(summary_est)
  )

ggplot(edu_ca) +
  geom_sf(aes(fill = percent)) +
  scale_fill_viridis_c(option = "C") +
  labs(title = "Adult Percentage with at least a Bachelor's in CA (2020)",
       fill = "% with Degree") +
  theme_minimal()
````


For educational attainment, we use `B15003_022` through `B15003_025` to sum all individuals with a bachelor’s degree or more, then divide by the total population (variable `B15003_001`). The `scale_fill_viridis_c()` function applies colorblind-friendly color scales to maps, using palettes like viridis, plasma, or magma.




```{r}
#| message: false
edu_ca <- edu_ca |>
  group_by(GEOID) |>
  summarize(
    percent = 100 * sum(estimate[variable != "B15003_001"]) / unique(summary_est)
  )

ggplot(edu_ca) +
  geom_sf(aes(fill = percent)) +
  scale_fill_viridis_c(option = "C") +
  labs(title = "Adult Percentage with at least a Bachelor's in CA (2020)",
       fill = "% with Degree") +
  theme_minimal()
```

###

`ggplot2` can handle spatial data directly using `geom_sf()`. Use `mutate()` to calculate percentages, and pipe that into `ggplot()` for a map.


### Exercise 9

In the R Terminal, run the following command to display the last chunk of your `.qmd` file

```
show_file("CaliforniaBachelors.qmd", chunk = "Last")
```

CP/CR

```{r california-bachelors-degree-9}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 15)
```

###

The variables in `get_acs()` like `"B19013_001"` are codes that represent specific census survey questions or data points.

 You can use the `load_variables()` function from tidycensus to look up what each code means and explore available variables for a given year and dataset. 
 
 This helps you find exactly the data you need without guessing variable names.


## California Median Age

### 

In this exercise, you will collect and plot median age data for all counties in California for the year 2020 using the **tidycensus** package.

Create a new Quarto Document and save it as `CaliforniaAge.qmd`. Add a YAML header at the top with a title (`"CaliforniaAge"`) and your name as author. Render the document using `Cmd/Ctrl + Shift + K`.




### Exercise 1

In your QMD, put `library(tidyverse)`, `library(tidycensus)`, and `library(sf)` in a new code chunk. Press Ctrl/Cmd + Shift + K to render the file

Notice that the file does not look good because the code is visible and there are annoying messages. To take care of this, add `#| message: false` to remove all the messages in this `setup` chunk. Also add the following to the YAML header to remove all code echoes from the HTML:

```
execute:
  echo: false
```

In the R Terminal, run:

```
show_file("CaliforniaAge.qmd", chunk = "Last")
```

CP/CR.

```{r california-median-age-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```

### 

Render again. Everything looks nice, albeit empty, because we have added code to make the file look better and more professional.

### Exercise 2

Place your cursor in the QMD file on the `library(tidyverse)` line. Use `Cmd/Ctrl + Enter` to execute that line.

Note that this causes `library(tidyverse)` to be copied down to the R Terminal and then executed. 

CP/CR.

```{r california-median-age-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

Working in the console like this is how professionals work!



### Exercise 3


Ask AI to generate R code that uses tidycensus data to get both the median age and populations for all counties in California for 2020 in a variable called `age_ca`

Put what it gives you in a new code chunk and do `Cmd/Ctrl + Enter` to send it to the console


CP/CR

```{r california-median-age-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```


### 

Our code:

````
age_ca <- get_acs(
  geography = "county",
  variables = c(median_age = "B01002_001", population = "B01003_001"),
  state = "CA",
  year = 2020,
)
````

AI is your best friend! Professionals use AI to generate code and plot. You however, need to be careful as it can and will generate much extra code that you may not need.


### Exercise 4

Using our code, add `geometry = FALSE`.

Take your code chunk and do `Cmd/Ctrl + Enter` to send it to the R Terminal.

In the R Terminal, run:

```         
show_file("CaliforniaAge.qmd", chunk = "Last")
```

CP/CR.

```{r california-median-age-4}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```

###

````
age_ca <- get_acs(
  geography = "county",
  variables = c(median_age = "B01002_001", population = "B01003_001"),
  state = "CA",
  year = 2020,
  geometry = FALSE
)
````


The American Community Survey (ACS) provides annual demographic, economic, and housing data based on samples, while the Decennial Census gives a complete count every 10 years.

### Exercise 5

Let's explore the data, run `glimpse(age_ca)` in the console.

CP/CR.

```{r california-median-age-5}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

Unlike the education data which had multiple variables per geography, requesting named variables like `median_age = "B01002_001"` creates a `variable` column with your custom names instead of Census codes. This makes the data easier to work with when pivoting to wide format later.

### Exercise 6

Let's get some quick stats for each column of the data. 

Run `summary(age_ca)` in the console.

CP/CR.


```{r california-median-age-6}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

`get_acs()` is your go-to for detailed annual demographic estimates from the American Community Survey. It returns both point estimates and margins of error (MOE) by default.

### Exercise 7

Send our code to the console from the previous exercise.

Now, type `age_ca` in the console.

CP/CR the first few lines.

```{r california-median-age-7}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

Helpful info for the next exercise: 
`geom_col()` from `ggplot2`: Creates bar charts where bar heights correspond to values in the data.

### Exercise 8

Now, copy/paste those few lines into your AI. Tell it to use the `age_ca` data and make a bar plot showing median age by county.

Put what it gives you in a new code chunk and do `Cmd/Ctrl + Enter` to send it to the console

CP/CR

```{r california-median-age-8}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```

###

Our Code:

````
age_ca_wide <- age_ca |>
  select(NAME, variable, estimate) |>
  pivot_wider(names_from = variable, values_from = estimate)


ggplot(age_ca_wide, aes(x = reorder(NAME, median_age), y = median_age)) +
  geom_col(fill = "#4daf4a") +
  coord_flip() 
````

Annotate charts with `geom_text()` or `labs()` to add clarity about what each axis, facet, or fill represents—especially useful for public-facing work.

### Exercise 9

Now we have a basic bar plot, but we have too much data points.

Tell your AI to filter the dataset to the 15 most populous counties and add informative labels and a title with `labs()`.

Put what it gives you in the  code chunk and do `Cmd/Ctrl + Enter` to send it to the console

CP/CR

```{r california-median-age-9}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```

###

Here is our code. It is okay if your code is different. That will happen when using AI!
Replace your code with what it gave you using this code: 

````
age_ca_wide <- age_ca |>
  select(NAME, variable, estimate) |>
  pivot_wider(names_from = variable, values_from = estimate)

largest_ca <- age_ca_wide |>
  arrange(desc(population)) |>
  slice_head(n = 15)

ggplot(largest_ca, aes(x = reorder(NAME, median_age), y = median_age)) +
  geom_col(fill = "#4daf4a") +
  coord_flip() +
  labs(
    title = "2020 California Counties Median Age",
    x = "County",
    y = "Median Age",
    caption = "Source: ACS 5-Year Estimates via tidycensus"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.text.y = element_text(size = 10)
  )
````

`theme_minimal()` from `ggplot2`:
Applies a clean, minimal theme to the plot for better readability.

```{r}
#| message: false
age_ca_wide <- age_ca |>
  select(NAME, variable, estimate) |>
  pivot_wider(names_from = variable, values_from = estimate)

largest_ca <- age_ca_wide |>
  arrange(desc(population)) |>
  slice_head(n = 15)

ggplot(largest_ca, aes(x = reorder(NAME, median_age), y = median_age)) +
  geom_col(fill = "#4daf4a") +
  coord_flip() +
  labs(
    title = "2020 California Counties Median Age",
    x = "County",
    y = "Median Age",
    caption = "Source: ACS 5-Year Estimates via tidycensus"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.text.y = element_text(size = 10)
  )
```


### Exercise 10

In the R Terminal, run the following command to display the last chunk of your `.qmd` file.

```
show_file("CaliforniaAge.qmd", chunk = "Last")
```

CP/CR

```{r california-median-age-10}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 15)
```

###

The `show_file()` function from tutorial.helpers is a convenient way to check the contents of files without leaving R. It helps confirm that your edits were saved properly.


## Summary

###

This tutorial covered an overview of [Analyzing US Census Data](https://walker-data.com/census-r/index.html) by Kyle Walker. You learned about using the **tidycensus** package for collecting, interacting, and plotting US Census data. You mainly focused on collecting data from the Decennial Census and the American Community Survey (ACS).

### Exercise 1

`Cmd/Ctrl + Shift + K` to ensure that all three of your QMDs render without errors. The resulting HTML pages should show your choropleth maps and bar chart.

In the R Terminal, run:

```
show_file("CaliforniaAge.qmd")
```

CP/CR.

```{r summary-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 30)
```

###

ACS data includes both an `estimate` column (the survey’s best guess) and a `moe` column (margin of error). The `geography` parameter in `get_acs()` allows data to be aggregated to common Census units — county, tract, state, and more. **tidycensus** accepts state names (e.g. `"Texas"`), postal codes (e.g. `"TX"`), and FIPS codes (e.g. `"48"`), so you can use whichever format you know.

### Exercise 2

Publish your rendered QMD to GitHub Pages. In a Bash terminal, run:

````
quarto publish gh-pages TexasIncome.qmd
````

If the bash terminal is still running from rendering, stop it first using `Ctrl/Cmd + C`.

Copy/paste the resulting URL below.

```{r summary-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 1)
```

###

Publishing the QMD creates a public URL anyone can use to view your choropleth map.

### Exercise 3

Commit and push all your files. Copy/paste the URL to your GitHub repo.

```{r summary-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

GitHub repositories are where people can view the code you used to create your project.

```{r download-answers, child = system.file("child_documents/download_answers.Rmd", package = "tutorial.helpers")}
```
