---
title: "AI and Git"
author: David Kane
tutorial:
  id: "07-ai-2"
output:
  learnr::tutorial:
    df_print: default
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: "A return to the Gemini CLI after learning Git. Sets up a fresh Quarto project with Gemini, then uses the agent to explain and run a full Git workflow --- status, staging, committing, and pushing --- and to publish to GitHub Pages."
---

```{r setup, include = FALSE}
library(learnr)
library(tutorial.helpers)
library(knitr)

knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(out.width = '90%')
options(tutorial.exercise.timelimit = 60,
        tutorial.storage = "local")
```

```{r info-section, child = system.file("child_documents/info_section.Rmd", package = "tutorial.helpers")}
```

## Introduction
###

This is the second tutorial about the [Gemini CLI](https://github.com/google-gemini/gemini-cli). In the **AI Introduction** tutorial you met Gemini as an *agent* and used it to create and render a Quarto analysis. Now that you have learned Git, you will use Gemini to handle the Git workflow --- and, just as importantly, to *explain* what it is doing, so you can check the agent's work against the Git you already know how to do by hand.

First, set up a fresh project for Gemini to work in.

### Exercise 1

You must do this tutorial in a Codespace created from a GitHub repo named `07-ai-2`. If you haven't already done this, follow these steps:

1. Create a GitHub Repo: go to the [codespace-starter repo](https://github.com/ppbds-student/codespace-starter). Click the green "Use this Template" drop-down button and select "Create a new repository". Name the repo `07-ai-2`. Then click "Create repository".

2. Create a Codespace from `07-ai-2`: from this repo's page, click the green "Code" drop down button, select the "Codespaces" tab, and click "Create Codespace on main".

Copy and paste the URL for your `07-ai-2` repository.

```{r introduction-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

Your answer should look something like:

````
https://github.com/ppbds-student/07-ai-2
````

###

The `codespace-starter` repo lists the Gemini CLI in its `.devcontainer/devcontainer.json`, so `gemini` is ready to run the moment the Codespace opens --- nothing to install.

### Exercise 2

Start Gemini by running `gemini` in the bash Terminal. As in the AI Introduction tutorial, press `y` to connect the Codespace and `y` to trust the directory. Because this is a new Codespace, you must authenticate with your Google account again: open the URL it prints, sign in, and paste the authorization code back into the terminal.

Once you reach the `>` prompt, CP/CR the first several lines of output.

```{r introduction-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 8)
```

###

Each Codespace is a fresh machine, so the OAuth token from your last Codespace does not carry over --- you sign in once per Codespace. This is the same authentication flow you walked through in the AI Introduction tutorial.

### Exercise 3

Recreate the analysis from the first tutorial. At the `>` prompt, ask Gemini to build it. Type something like:

```
Create an analysis.qmd for a Quarto project. Give it a YAML header with title "Exploring Storms", my name as author, and execute: echo: false. Add a setup chunk that loads the tidyverse with suppressPackageStartupMessages(). Then add a chunk that plots the number of named storms per year using the storms dataset from dplyr. Render it when you are done.
```

Press `y` to let Gemini write `analysis.qmd` and run `quarto render`. CP/CR Gemini's summary of what it created.

```{r introduction-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 12)
```

###

Notice that you described the *result* you wanted --- a plot of storms per year --- and let Gemini choose the functions. That is the core skill: steer the agent toward a goal, then check what it produced.

### Exercise 4

Now have Gemini add a `.gitignore`. Type something like:

```
Create a .gitignore file for a Quarto R project. Include patterns for Quarto cache and build directories, but do NOT ignore .html files.
```

Press `y` to authorize the write. Then type `/quit` to leave Gemini. CP/CR.

```{r introduction-4}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 12)
```

###

You kept `.html` out of `.gitignore` on purpose. Publishing to GitHub Pages needs the rendered `analysis.html`, so it must be committed --- which is exactly what you will do, with Gemini's help, in a moment.

### Exercise 5

Confirm the project is ready. In the bash Terminal, run `ls -a`. CP/CR.

```{r introduction-5}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

Your answer should look something like:

````
@ppbds-student ➜ /workspaces/07-ai-2 (main) $ ls -a
.  ..  .devcontainer  .git  .gitignore  analysis.html  analysis.qmd
@ppbds-student ➜ /workspaces/07-ai-2 (main) $
````

You now have the three files the next section will put under version control: the source `analysis.qmd`, the rendered `analysis.html`, and the `.gitignore`. From here, Gemini takes over Git.

## Git with Gemini
###

Now that the project has real files, let's save the work with Git. Rather than just telling you the commands, we'll use Gemini to explain the Git workflow. Using AI to understand what you are doing --- not just to do it --- is a powerful learning strategy.

### Exercise 1

Ask Gemini to check the Git status and explain what the output means:

```
Run git status and explain what each part of the output means.
```

Press `y`, 1, or hit Enter on allow once to allow the shell command. CP/CR the full response including Gemini's explanation.

```{r git-with-gemini-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 18)
```

###

Your answer should look something like:

````
> Run git status and explain what each part of the output means.

╔ Tool Use: run_shell ══════════════════╗
║  Command: git status                  ║
╚═══════════════════════════════════════╝

Allow this tool call? [y/n/a]: y

Output:
On branch main
Your branch is up to date with 'origin/main'.

Untracked files:
  (use "git add <file>..." to include in what will be committed)
        .gitignore
        analysis.html
        analysis.qmd

nothing added to commit but untracked files present

Here's what this means:
- "On branch main" — you are working on the main branch.
- "Your branch is up to date" — your local branch matches the GitHub copy.
- "Untracked files" — Git has noticed these files but is not monitoring them yet. They will not be included in any commit until you explicitly add them with `git add`.

>
````

###

`git status` shows your branch, **staged** changes (added to the next commit), and **untracked** files (noticed by Git but not yet included). Smart commit in `codespace-starter` handles staging automatically --- just press the "Commit" button.

You may notice that `analysis.html` does not appear in your `git status` output. This can happen if the `.gitignore` Gemini created includes `*.html`, which tells Git to ignore all HTML files. If this happens, you can use `git add -f analysis.html` to force Git to track it anyway.

### Exercise 2

Ask Gemini to explain the difference between `git add` and `git commit`, and to provide the exact command to commit all three files with a good message:

```
Explain the difference between git add and git commit. Then give me the exact command to commit .gitignore, analysis.qmd, and analysis.html with a descriptive message.
```

CP/CR Gemini's full response.

```{r git-with-gemini-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 20)
```

###

Gemini should explain that `git add` moves files into the **staging area**, a holding zone where you prepare which changes to include before committing. `git commit` then takes everything in the staging area and permanently records it in the project's history. A typical response will include commands like:

```
git add .gitignore analysis.qmd analysis.html
git commit -m "initial version with analysis and gitignore"
```

###

The two-step design lets you choose exactly which files go into each commit --- useful when you want to record related changes separately with different messages. In `codespace-starter`, the Source Control sidebar handles both steps at once, but knowing they are separate matters when collaborating on larger projects.

### Exercise 3

Open the Source Control sidebar (the branch icon in the Activity Bar). You should see `.gitignore`, `analysis.qmd`, and `analysis.html` listed as changes. Type a commit message in the message box at the top, then click the **Commit** button. Because smart commit is enabled in the devcontainer, clicking Commit will stage and commit all changes in one step --- no `git add` required.

Then, in a separate bash Terminal, run:

```
git log --oneline
```

CP/CR.

```{r git-with-gemini-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```

###

Your answer should look something like:

````
@ppbds-student ➜ /workspaces/07-ai-2 (main) $ git log --oneline
a3b8c21 (HEAD -> main) initial version with analysis and gitignore
e3b01ee Initial commit
@ppbds-student ➜ /workspaces/07-ai-2 (main) $
````

The `(HEAD -> main)` indicator confirms your local branch is current, but the missing `origin/main` means this commit hasn't been pushed to GitHub yet. Pushing is what sends it there and makes it visible to others.

### Exercise 4

Ask Gemini to improve the storms-per-year chart. Type something like:

```
Update the storms-per-year chart in analysis.qmd to be a stacked area chart showing the count of each storm type per year. Filter to just tropical depressions, tropical storms, and hurricanes. Use a color scale that goes from light blue for tropical depressions through orange for tropical storms to dark red for hurricanes. Re-render the file when done.
```

Press `y` to authorize each tool call. Once the Live Server tab has refreshed and you are happy with the result, ask Gemini to handle the entire Git workflow:

```
Please commit analysis.qmd and analysis.html with a good message, and push to GitHub.
```

Press `y` to authorize each tool call as Gemini works through the steps. CP/CR the complete output.

```{r git-with-gemini-4}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 20)
```

###

Your output should look something like:

````
> Please commit analysis.qmd and analysis.html with a good message, and push to GitHub.

╔ Tool Use: run_shell ═══════════════════════════════════════╗
║  Command: git add analysis.qmd analysis.html               ║
╚════════════════════════════════════════════════════════════╝
Allow this tool call? [y/n/a]: y

╔ Tool Use: run_shell ══════════════════════════════════════════════╗
║  Command: git commit -m "convert to stacked area chart by storm type" ║
╚══════════════════════════════════════════════════════════════════╝
Allow this tool call? [y/n/a]: y

╔ Tool Use: run_shell ════════════════════════╗
║  Command: git push origin main              ║
╚═════════════════════════════════════════════╝
Allow this tool call? [y/n/a]: y

✔ Done! Your changes have been committed and pushed to GitHub.

>
````

###

Gemini issued three separate tool calls --- `git add`, `git commit`, and `git push` --- mapping directly to the three stages of staging, recording, and sharing. In Exercise 3 you ran these yourself using the Source Control sidebar; here you delegated them to gemini.

### Exercise 5

Still in Gemini, type `/stats session` and then `/stats model`. CP/CR both outputs.

```{r git-with-gemini-5}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 18)
```

###

Your output should look something like:

````
> /stats session

Session Statistics:
  Messages:           14
  Tool calls made:    10
  Tool calls allowed: 10
  Input tokens:       ~18,400
  Output tokens:      ~5,100
  Total tokens:       ~23,500

> /stats model

Model: gemini-2.5-pro
  Context window:    1,000,000 tokens
  Free tier limit:   1,000,000 tokens/day
  Used today:        ~23,500 tokens
  Remaining:         ~976,500 tokens

>
````

###

AI models have **context windows** --- a limit on how much text they can process in one session. Gemini's are very large (up to 1 million tokens), but `/stats` lets you track usage in real time; on the free tier, your daily limit resets every day.


## Summary
###

This tutorial introduced the [Gemini CLI](https://github.com/google-gemini/gemini-cli) as a tool for AI-assisted data science in GitHub Codespaces. You authenticated with your Google account, used Gemini to create and render a Quarto analysis file, and delegated a complete Git workflow --- staging, committing, and pushing --- to the CLI.

### Exercise 1

Quit Gemini with `/quit` if you haven't already. From the bash Terminal, run `quarto render analysis.qmd` to regenerate the HTML. Then, in the R Terminal, run:

```
tutorial.helpers::show_file("analysis.qmd")
```

CP/CR.

```{r summary-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 30)
```

###

Your answer should show the full contents of `analysis.qmd`, including a YAML header with `execute: echo: false`, a setup chunk using `suppressPackageStartupMessages(library(tidyverse))`, and at least two ggplot2 code chunks creating plots.

### Exercise 2

Publish your rendered QMD to GitHub Pages. In the bash Terminal, run:

```
quarto publish gh-pages analysis.qmd
```

Respond with `Y` when prompted. Copy/paste the resulting URL below.

```{r summary-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

Your URL should look something like:

```
https://ppbds-student.github.io/07-ai-2/
```

GitHub Pages takes a minute or two to deploy after `quarto publish` runs. If the link doesn't load immediately, wait and refresh.

### Exercise 3

Commit and push any remaining uncommitted files. Copy/paste the URL to your GitHub repo.

```{r summary-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

Other AI coding tools include [Claude Code](https://claude.ai/claude-code) (Anthropic), [GitHub Copilot](https://github.com/features/copilot) (Microsoft), and [Cursor](https://cursor.sh/) (Anysphere). The workflow is the same across all of them: describe what you want, review the result, and iterate.

```{r download-answers, child = system.file("child_documents/download_answers.Rmd", package = "tutorial.helpers")}
```
