---
title: First Tutorial for R4DS
author: David Kane
tutorial:
  id: r4ds-1
output:
  learnr::tutorial:
    df_print: default
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: 'Introduces data analysis with music and chart data. Covers importing CSV files, transforming and tidying tabular data, and building visualizations using the Billboard Hot 100 and Million Song datasets.'
---

```{r setup, include = FALSE}
library(learnr)
library(tutorial.helpers)
library(tidyverse)
library(quarto) # Total hack to shut up renv! Really need to put this in the package help file.

knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(out.width = '90%')
options(tutorial.exercise.timelimit = 60,
        tutorial.storage = "local")

billboard_clean <- billboard |>
  filter(wk1 < 101) |>
  arrange(wk1) |>
  mutate(improvement = wk1 - wk2, .after = wk2) |>
  select(artist:improvement) |>
  rename(week1 = wk1, week2 = wk2)

billboard_longer <- billboard |>
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE
  ) |>
  mutate(week = parse_number(week))
```

```{r info-section, child = system.file("child_documents/info_section.Rmd", package = "tutorial.helpers")}
```

## Introduction
###

<!-- AR:  -->

This tutorial analyzes two real music datasets: the **Billboard Hot 100**, which tracks weekly chart rankings for popular songs, and the **Million Song Dataset**, a sample of 10,000 songs with audio features and metadata.

You will use **[readr](https://readr.tidyverse.org/)**, **[dplyr](https://dplyr.tidyverse.org/)**, and **[ggplot2](https://ggplot2.tidyverse.org/)** to import, transform, and visualize music data. Key functions include [`read_csv()`](https://readr.tidyverse.org/reference/read_delim.html), [`filter()`](https://dplyr.tidyverse.org/reference/filter.html), [`ggplot()`](https://ggplot2.tidyverse.org/reference/ggplot.html), and [`pivot_longer()`](https://tidyr.tidyverse.org/reference/pivot_longer.html).

We recommend using an AI agent that works directly within your development environment, such as [Gemini CLI](https://github.com/google-gemini/gemini-cli) or [Claude Code](https://claude.ai/code), which can read and edit your files without copy/pasting. You may also use a chat-based AI (such as ChatGPT or Claude on the web), though this will require manually copying code into your Quarto document. Instructions throughout these tutorials are written assuming you are using an in-environment AI agent, but the choice is yours.

### Exercise 1

<!-- AR: Add link. -->

Create a new repo using []`codespace-starter`]() as a template and name it `r4ds-1`.

If you haven't already, open the `r4ds-1` repo on GitHub and create a Codespace from it.

You need two R Terminals: one for running this tutorial and one for your exercises.

Create a new Quarto Document and save it as `analysis.qmd`. Add a YAML header at the top with a title (`"Analyzing the Billboard 100"`) and your name as author. In a bash terminal, run:

```
quarto render analysis.qmd
```

Open `analysis.html` in a browser by right-clicking it in the File Explorer and selecting **Open with Live Server**. Keep this tab open — it will refresh automatically whenever you re-render.

Create a `.gitignore` file with `analysis_files` on the first line and then a blank line. Save and push.

In the R Terminal, run:

```
show_file(".gitignore")
```

If that fails, it is probably because you have not yet loaded `library(tutorial.helpers)` in the R Terminal.

CP/CR.

```{r introduction-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

<!-- AR: Show example show file text. CLAUDE should know that every exercise has something with which students can verify their submissions. 

<!-- Can it appear after hitting submit, or is a continue button necessary? -->
-->

###

<!-- AR: We don't want to have any default knowledge drops. Instead, include something from the bloated introduction. Background -->

Professionals keep their data science work in the cloud because laptops fail.

### Exercise 2

In your QMD, put `library(tidyverse)` in a new code chunk. In a bash terminal, run `quarto render analysis.qmd`. Check your Live Server tab — it should refresh automatically.

Notice that the file does not look good because the code is visible and there are annoying messages. To take care of this, add `#| message: false` to remove all the messages in this chunk. For historical reasons, the first chunk in a Quarto document is often referred to as the `setup` chunk. Also, add the following to the YAML header to remove all code echoes from the HTML:

```
execute:
  echo: false
```

In a bash terminal, run `quarto render analysis.qmd` again. Check your Live Server tab. Only the title and author should appear in the HTML.

In the R Terminal, run:

```
show_file("analysis.qmd", chunk = "Last")
```

CP/CR.

```{r introduction-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```

###

You may get a warning like:

```
Warning message:
In readLines(path) : incomplete final line found on 'analysis.qmd'
```

This means that `analysis.qmd` lacks an empty last line. Add one. Text files should usually have a blank line at the end.

### Exercise 3

In a bash terminal, run `quarto render analysis.qmd`. Confirm that the Live Server tab auto-refreshes.

CP/CR.

<!-- AR: What specifically, should be pasted? -->

```{r introduction-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

Note that **ggplot2** is loaded with **tidyverse**.

R has several systems for making graphs, but [**ggplot2**](https://ggplot2.tidyverse.org/index.html) is the best. **ggplot2** implements the *grammar of graphics*, a coherent system for describing and building graphs.

### Exercise 4

From the R Terminal, run these three commands:

<!-- AR: Do something so these are displayed on 3 separate lines in the rendered tutorial. -->

`getwd()`
`dir.create("data")`
`list.files()`

These commands organize your project directories: `getwd()` shows your working directory (WD), which is your current location. `dir.create("data")` creates a data folder in your WD and `list.files()` displays files in WD when parentheses are empty.

CP/CR.

```{r introduction-4}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

Your answer should look something like this, although your path will be different.

````
> getwd()
[1] "/workspaces/r4ds-1"
> dir.create("data")
> list.files()
 [1] "analysis_files" "analysis.html"  "analysis.qmd"   "data"
>
````

<!-- AR: Needs a knowledge drop here. One idea: You could ask Cluade to do this for you-->


## Billboard 100
###

The [`billboard`](https://tidyr.tidyverse.org/reference/billboard.html) tibble is a dataset in the [**tidyr**](https://tidyr.tidyverse.org/index.html) package. `billboard` contains data from 317 songs and their weekly rankings in the Billboard top 100 for 76 weeks after they were released.

###

```{r billboard-data}
billboard
```

### Exercise 1

Before plotting, inspect the data. Ask AI to add a code chunk to your QMD that prints `billboard`. Render and check your Live Server tab.

Copy and paste the table from the HTML.

```{r billboard-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 4)
```

###

```{r billboard-1-test}
billboard
```

`billboard` has 317 rows (one per song) and 79 columns: `artist`, `track`, `date.entered`, and 76 weekly ranking columns (`wk1`–`wk76`). Missing values mean the song dropped off the chart before that week. This "wide" format is convenient for data entry but awkward for analysis — you'll tidy it later with `pivot_longer()`.

### Exercise 2

<!-- AR: Going way too fast! Show me a histogram of week 1. Show me the top 5 artists. Glimpse, summary... add this guidance to CLAUDE.md

Exploration: 5-6 question
- univariate

What are the variables?

Ask Ai?
- are there any songs that enter, exit, and enter again
- what are the key variables and summarize
- explain this data set to me

hist of week 1 shows there are no na's week 6 has more
week 1 ranks are lower than later

hist of week 4

Plots
- answer interesting questions
- think about the data
- make one where each line is a song's trajectory in the billboard

example commands
sum(billboards$wk1)

 -->

Loading the **tidyverse** package automatically loaded the packages which are part of the "core" *Tidyverse*, including **tidyr**, **ggplot2**, and **dplyr**.

Ask AI to create a scatter plot of `wk1` vs `wk2` from `billboard` using `geom_point()` and `geom_smooth()`. Include a proper title, subtitle, caption, and axis labels in `labs()`. Add this code to a new code chunk in your QMD. In a bash terminal, run `quarto render analysis.qmd`. Check your Live Server tab.

In the R Terminal, run:

```
show_file("analysis.qmd", chunk = "Last")
```

CP/CR.

```{r billboard-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

```{r billboard-2-test}
ggplot(billboard, aes(x = wk1, y = wk2)) +
  geom_point() +
  geom_smooth() +
  labs(
    title = "Billboard Rankings: Week 1 vs. Week 2",
    subtitle = "Shows how a song's debut ranking relates to its second week performance",
    caption = "Source: Billboard dataset (2000)",
    x = "Week 1 Rank",
    y = "Week 2 Rank"
  )
```

A trend line shows the general pattern in your data. The shaded area around the line represents uncertainty — a wider band means less certainty. **ggplot2** supports both linear and curved (LOESS) trend lines.

### Exercise 3

Create A new tibble with columns for artist, track, week 1 rank, week 2 rank, and a new `improvement` column (week 1 rank minus week 2 rank) and show me a histogram of improvement.

In the R Terminal, run:

```
show_file("analysis.qmd", chunk = "Last")
```

CP/CR.

```{r billboard-3}
question_text(NULL,
    answer(NULL, correct = TRUE),
    allow_retry = TRUE,
    try_again_button = "Edit Answer",
    incorrect = NULL,
    rows = 6)
```

###

Our result:

```{r billboard-3-test}
billboard_clean
```

<!-- AR: focus on packages -->

`filter()` selects rows by condition, `arrange()` reorders them, `mutate()` creates new columns, `select()` keeps specified columns, and `rename()` changes column names — all without modifying the original tibble.

### Exercise 4

<!-- AR: Wait until some more work has been done to the data set. Want this to actually save time after being done. Probably after making a picture. -->

Within the latest code chunk, add the option: `#| cache: true`. Assign the result of the pipe to `billboard_clean`. If you have not done so already, add `analysis_cache` to your `.gitignore`.

In a bash terminal, run `quarto render analysis.qmd`. This first render will take some time — by including `#| cache: true`, Quarto saves the result to disk so the next render loads from cache instead of re-running. Check your Live Server tab.

In the R Terminal, run:

```
show_file("analysis.qmd", chunk = "Last")
```

CP/CR.

```{r billboard-4}
question_text(NULL,
    answer(NULL, correct = TRUE),
    allow_retry = TRUE,
    try_again_button = "Edit Answer",
    incorrect = NULL,
    rows = 8)
```

###

<!-- AR: Need new drop.

Drops can be about infrastructure as well?

 -->

### Exercise 5

Ask AI to pipe `billboard_clean` into a histogram of `improvement` with a title, subtitle, and caption. Add this code to a new code chunk in your QMD. In a bash terminal, run `quarto render analysis.qmd`. Check your Live Server tab.

In the R Terminal, run:

```
show_file("analysis.qmd", chunk = "Last")
```

CP/CR.

```{r billboard-5}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 12)
```

###

```{r billboard-5-test}
ggplot(billboard_clean, aes(x = improvement)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") +
  labs(
	title = "Distribution of Billboard Top 100 Song Ranking Improvements From Week 1 to Week 2",
	subtitle = "Almost no songs decreased in rank between weeks 1 and 2",
	x = "Song Rank Improvement",
	y = "Count",
	caption = "Source: Rankings for songs that appeared in the Billboard top 100 in the year 2000"
  ) +
  theme(
    plot.title = element_text(size = 7.5, face = "bold"),
    plot.subtitle = element_text(size = 8)
  )
```

A histogram divides the x-axis into equally spaced bins and uses the bar height to display observation counts. Many `theme_*()` functions exist within **ggplot2** that create different appearance themes for plots.

<!-- AR: Cache and commit/push exercise(s) here! Only cache when you are done iterating a visualization and you don't want to re-run the code again. Adjust .gitignore for the cache directory and commit and push here too. -->

## Music
###

The **music** CSV file comes from the [**Million Song Dataset**](https://corgis-edu.github.io/corgis/csv/music/), which used a company called the Echo Nest to derive data points about one million popular contemporary songs. **Music** only contains 10,000 rows of data from the **Million Song Dataset**.

Key functions you'll learn: [`read_csv()`](https://readr.tidyverse.org/reference/read_delim.html), [`pivot_longer()`](https://tidyr.tidyverse.org/reference/pivot_longer.html), [`parse_number()`](https://readr.tidyverse.org/reference/parse_number.html).

### Exercise 1

<!-- AR: For all future tutorials, just tell AI to download from the URL into the data directory -->

We begin by downloading our CSV file `music.csv` to the `data` directory directly from the internet using `download.file()`.

In the R Terminal, run:

```
download.file(
  "https://github.com/PPBDS/vscode.tutorials/raw/refs/heads/main/inst/extdata/r4ds-1/music.csv",
  destfile = "data/music.csv",
  mode = "wb"
)
```

CP/CR.

```{r music-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

"CSV" stands for **C**omma-**S**eparated **V**alues, meaning that the variable names and data values are separated by commas in the file. `download.file()` saves a file from a URL to a local path. The `mode = "wb"` argument is required for binary files — without it, text-mode line-ending conversion can corrupt the file.

### Exercise 2



Ask AI to read `data/music.csv` using `read_csv()`. Don't assign the result to a variable. Add this code to a new code chunk in your QMD. In a bash terminal, run `quarto render analysis.qmd`.

Copy and paste the result from the HTML.

```{r music-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```{r music-2-test}
read_csv("../../extdata/r4ds-1/music.csv", show_col_types = FALSE)
```


<!-- AR: Say something that motivates the next AI question. 

How many artists are there? What are the most common artist ids? Make me a map of the artist locations using the latitude and longitude. 

Whenever we get a data set poke around and explore it

Ask what artist familiarity is? Copy and paste the answer. Can use documentation website of AI. Then show the proper definition from original docs

Data file is only CSV without a README - should course materials include the README as well?

While making plots, use AI to help me make a cool plot. I should be playing around with these data. Which variables do you think are most useful? What are the most noteable features?
-->

### Exercise 3

Ask AI to create a tidy version of `billboard` using `pivot_longer()` to reshape the weekly ranking columns into two columns: `week` and `rank`. Tell AI to drop rows with missing rankings and convert `week` from a character string to a number. Add this code to a new chunk in your QMD and render.

In the R Terminal, run:

```
show_file("analysis.qmd", chunk = "Last")
```

CP/CR.

```{r music-3}
question_text(NULL,
    answer(NULL, correct = TRUE),
    allow_retry = TRUE,
    try_again_button = "Edit Answer",
    incorrect = NULL,
    rows = 6)
```

###

Our result:

```{r music-3-test}
billboard_longer
```

[`pivot_longer()`](https://tidyr.tidyverse.org/reference/pivot_longer.html) "lengthens" data by turning column names into a variable (`names_to`) and values into another (`values_to`). `parse_number()` extracts the first number from a string, making `week` numeric for plotting.

### Exercise 4

Within the latest code chunk, add the option: `#| cache: true`. Assign the result of the pipe to `billboard_longer`. If you have not done so already, add `analysis_cache` to your `.gitignore`.

In a bash terminal, run `quarto render analysis.qmd`. This first render will take some time — by including `#| cache: true`, Quarto saves the result to disk so the next render loads from cache instead of re-running. Check your Live Server tab.

In the R Terminal, run:

```
show_file("analysis.qmd", chunk = "Last")
```

CP/CR.

```{r music-4}
question_text(NULL,
    answer(NULL, correct = TRUE),
    allow_retry = TRUE,
    try_again_button = "Edit Answer",
    incorrect = NULL,
    rows = 8)
```

###

There are three interrelated rules that make a dataset tidy: each variable is a column, each observation is a row, and each value is a cell. Most real-world data is untidy because it's often organized for data entry rather than analysis.

### Exercise 5

Ask AI to pipe `billboard_longer` into a plot showing the Billboard rank of each track over the weeks after they were released. Within `labs()`, add a proper title, subtitle, and caption. Don't assign the plot. Add this code to a new code chunk in your QMD. In a bash terminal, run `quarto render analysis.qmd`. Check your Live Server tab.

In the R Terminal, run:

```
show_file("analysis.qmd", chunk = "Last")
```

CP/CR.

```{r music-5}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 12)
```

###

```{r music-5-test}
billboard_longer |>
  ggplot(aes(x = week, y = rank, group = track)) +
  geom_line(alpha = 0.2) +
  scale_y_reverse() +
  labs(
    title = "Billboard Track Rankings Over Time",
	subtitle = "Most tracks rise in Billboard rankings in their first 10 weeks",
    x = "Weeks Since Release",
    y = "Billboard Rank",
	caption = "Source: billboard from tidyr"
  )
```

Most real analyses will require at least a little tidying. You'll begin by figuring out what the underlying variables and observations are, then pivot your data into a tidy form with variables in the columns and observations in the rows.

<!-- AR: This data is not reflecting the real world -->


## Summary
###

This tutorial analyzed two real music datasets: the **Billboard Hot 100** and a sample from the **Million Song Dataset**.

You used **[readr](https://readr.tidyverse.org/)**, **[dplyr](https://dplyr.tidyverse.org/)**, and **[ggplot2](https://ggplot2.tidyverse.org/)** to import, transform, and visualize music data. Key functions included [`read_csv()`](https://readr.tidyverse.org/reference/read_delim.html), [`filter()`](https://dplyr.tidyverse.org/reference/filter.html), [`ggplot()`](https://ggplot2.tidyverse.org/reference/ggplot.html), and [`pivot_longer()`](https://tidyr.tidyverse.org/reference/pivot_longer.html).

### Exercise 1

In a bash terminal, run `quarto render analysis.qmd` to ensure that everything works. Check your Live Server tab — the resulting HTML page should be attractive, showing clean versions of your plots.

In the R Terminal, run:

```
tutorial.helpers::show_file("analysis.qmd")
```

CP/CR.

```{r summary-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 30)
```

###

Tabular data is considered *tidy* if and only if it satisfies the following rules:

1.  Each variable is a column; each column is a variable.
2.  Each observation is a row; each row is an observation.
3.  Each value is a cell; each cell is a single value.

### Exercise 2

Publish your rendered QMD to GitHub Pages. In a bash terminal, run:

````
quarto publish gh-pages analysis.qmd
````

If the bash terminal is still running from rendering, stop it first using `Ctrl/Cmd + C`.

Copy/paste the resulting URL below.

```{r summary-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 1)
```

###

Publishing the QMD creates a public URL anyone can use to view the rendered project.

### Exercise 3

Commit and push all your files. Copy/paste the URL to your Github repo.

```{r summary-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

GitHub repositories are where people can view the code you used to create your project.

###

The functions you used today — `filter()`, `arrange()`, `mutate()`, `select()`, `rename()`, `pivot_longer()`, and `read_csv()` — are covered in depth in [*R for Data Science (2e)*](https://r4ds.hadley.nz/) by Hadley Wickham, Mine Çetinkaya-Rundel, and Garrett Grolemund. The chapters on [data visualization](https://r4ds.hadley.nz/data-visualize.html), [data transformation](https://r4ds.hadley.nz/data-transform.html), [data tidying](https://r4ds.hadley.nz/data-tidy.html), and [data import](https://r4ds.hadley.nz/data-import.html) are worth reading to deepen your understanding of what to ask AI for and how to evaluate its output.

```{r download-answers, child = system.file("child_documents/download_answers.Rmd", package = "tutorial.helpers")}
```
