---
title: First Tutorial for R4DS
author: Luke Li and David Kane
tutorial:
  id: r4ds-1
output:
  learnr::tutorial:
    df_print: default
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: 'This tutorial for "R for Data Science" covers Chapter 1: Data visualization, Chapter 3: Data transformation, Chapter 5: Data tidying, and Chapter 7: Data import. Exercises build ggplot2 plots and dplyr/tidyr pipelines on real datasets.'
---

```{r setup, include = FALSE}
library(learnr)
library(tutorial.helpers)
library(tidyverse)
library(quarto) # Total hack to shut up renv! Really need to put this in the package help file.

knitr::opts_chunk$set(echo = FALSE)
options(tutorial.exercise.timelimit = 60,
        tutorial.storage = "local")

line_graph <- billboard |>
  filter(track %in% c("Dancing Queen", "Loser", "Try Again")) |>
  pivot_longer(
    cols = wk1:wk5,
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE
  ) |>
  ggplot(aes(x = week, y = rank, color = track, group = track)) +
  geom_line() +
  facet_wrap(~ track)

billboard_clean <- billboard |>
  filter(wk1 < 101) |>
  arrange(wk1) |>
  mutate(improvement = wk1 - wk2, .after = wk2) |>
  select(artist:improvement) |>
  rename(week1 = wk1, week2 = wk2)

music_1 <- read_csv(
   "data/music.csv",
   col_types = cols_only(
     artist.familiarity = col_character(),
 artist.hotttnesss = col_character(),
 artist.id = col_character()
   )
)

df <- read_csv(
   "data/music.csv",
   col_types = cols(
     artist.familiarity = col_double()
   )
)

billboard_longer <- billboard |>
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE
  ) |>
  mutate(week = parse_number(week))
```

```{r info-section, child = system.file("child_documents/info_section.Rmd", package = "tutorial.helpers")}
```

## Introduction
###

This section covers key concepts from [Chapter 1: Data visualization](https://r4ds.hadley.nz/data-visualize.html), [Chapter 3: Data transformation](https://r4ds.hadley.nz/data-transform.html), [Chapter 5: Data tidying](https://r4ds.hadley.nz/data-tidy.html), and [Chapter 7: Data import](https://r4ds.hadley.nz/data-import.html) from [*R for Data Science (2e)*](https://r4ds.hadley.nz/) by Hadley Wickham, Mine Çetinkaya-Rundel, and Garrett Grolemund.

You will learn about importing, organizing, and visualizing data using core packages such as **[readr](https://readr.tidyverse.org/)**, **[dplyr](https://dplyr.tidyverse.org/)**, and **[ggplot2](https://ggplot2.tidyverse.org/)**. Important functions include: [`read_csv()`](https://readr.tidyverse.org/reference/read_delim.html), [`filter()`](https://dplyr.tidyverse.org/reference/filter.html), [`ggplot()`](https://ggplot2.tidyverse.org/reference/ggplot.html), and [`pivot_longer()`](https://tidyr.tidyverse.org/reference/pivot_longer.html).

Throughout these tutorials, you will use AI to help write and debug code. We recommend using an AI agent that works directly within your development environment, such as [Gemini CLI](https://github.com/google-gemini/gemini-cli) or [Claude Code](https://claude.ai/code), which can read and edit your files without copy/pasting. You may also use a chat-based AI (such as ChatGPT or Claude on the web), though this will require manually copying code into your Quarto document. Instructions throughout these tutorials are written assuming you are using an in-environment AI agent, but the choice is yours.

### Exercise 1

Create a new repo using `codespace-starter` as a template and name it `r4ds-1`.

If you haven't already, open the `r4ds-1` repo on GitHub and create a Codespace from it.

You need two R Terminals: one for running this tutorial and one for your exercises.

Create a new Quarto Document and save it as `analysis.qmd`. Add a YAML header at the top with a title (`"Analyzing the Billboard 100"`) and your name as author. In a bash terminal, run:

```
quarto render analysis.qmd
```

Open `analysis.html` in a browser by right-clicking it in the File Explorer and selecting **Open with Live Server**. Keep this tab open — it will refresh automatically whenever you re-render.

Create a `.gitignore` file with `analysis_files` on the first line and then a blank line. Save and push.

In the R Terminal, run:

```
show_file(".gitignore")
```

If that fails, it is probably because you have not yet loaded `library(tutorial.helpers)` in the R Terminal.

CP/CR.

```{r introduction-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

Professionals keep their data science work in the cloud because laptops fail.

### Exercise 2

In your QMD, put `library(tidyverse)` in a new code chunk. In a bash terminal, run `quarto render analysis.qmd`. Check your Live Server tab — it should refresh automatically.

Notice that the file does not look good because the code is visible and there are annoying messages. To take care of this, add `#| message: false` to remove all the messages in this chunk. For historical reasons, the first chunk in a Quarto document is often referred to as the `setup` chunk. Also, add the following to the YAML header to remove all code echoes from the HTML:

```
execute:
  echo: false
```

In a bash terminal, run `quarto render analysis.qmd` again. Check your Live Server tab. Only the title and author should appear in the HTML.

In the R Terminal, run:

```
show_file("analysis.qmd", chunk = "Last")
```

CP/CR.

```{r introduction-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```

###

You may get a warning like:

```
Warning message:
In readLines(path) : incomplete final line found on 'analysis.qmd'
```

This means that `analysis.qmd` lacks an empty last line. Add one. Text files should usually have a blank line at the end.

### Exercise 3

In a bash terminal, run `quarto render analysis.qmd`. Check your Live Server tab.

Note how this command causes `library(tidyverse)` to be copied down to the R Terminal and then executed.

CP/CR.

```{r introduction-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 8)
```

###

You should practice moving from QMD World to R Terminal World and back.

Note that **ggplot2** is loaded with **tidyverse**.

R has several systems for making graphs, but [**ggplot2**](https://ggplot2.tidyverse.org/index.html) is the best. **ggplot2** implements the *grammar of graphics*, a coherent system for describing and building graphs.

### Exercise 4

From the R Terminal, run these three commands:

`getwd()`
`dir.create("data")`
`list.files()`

These commands organize your project directories: `getwd()` shows your working directory (WD), which is your current location. `dir.create("data")` creates a data folder in your WD and `list.files()` displays files in WD when parentheses are empty.

CP/CR.

```{r introduction-4}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

Your answer should look something like this, although your path will be different.

````
> getwd()
[1] "/workspaces/r4ds-1"
> dir.create("data")
> list.files()
 [1] "analysis_files" "analysis.html"  "analysis.qmd"   "data"
>
````


## Billboard 100
###

The [`billboard`](https://tidyr.tidyverse.org/reference/billboard.html) tibble is a dataset in the [**tidyr**](https://tidyr.tidyverse.org/index.html) package. `billboard` contains data from 317 songs and their weekly rankings in the Billboard top 100 for 76 weeks after they were released.

### Exercise 1

Loading the **tidyverse** package automatically loaded the packages which are part of the "core" *Tidyverse*, including **tidyr**, **ggplot2**, and **dplyr**.

Type `billboard` into the R Terminal.

CP/CR.

```{r billboard-100-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

```{r billboard-100-1-test}
billboard
```

###

A "variable" represents a quantity, quality, or property that can be measured or observed. Variables can be numeric or categorical, continuous or discrete, qualitative or quantitative.

**Tabular data** is data organized in a table. A table is a group of cells, organized in rows and columns.

### Exercise 2

Ask AI to create a scatter plot of `wk1` vs `wk2` from `billboard` using `geom_point()` and `geom_smooth()`. Include a proper title, subtitle, caption, and axis labels in `labs()`. Add this code to a new code chunk in your QMD. In a bash terminal, run `quarto render analysis.qmd`. Check your Live Server tab.

CP/CR.

```{r billboard-100-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

```{r billboard-100-2-test}
ggplot(billboard, aes(x = wk1, y = wk2)) +
  geom_point() +
  geom_smooth() +
  labs(
    title = "Billboard Rankings: Week 1 vs. Week 2",
    subtitle = "Shows how a song's debut ranking relates to its second week performance",
    caption = "Source: Billboard dataset (2000)",
    x = "Week 1 Rank",
    y = "Week 2 Rank"
  )
```

###

A trend line shows the general pattern in your data. The shaded area around the line represents uncertainty — a wider band means less certainty. **ggplot2** supports both linear and curved (LOESS) trend lines.

### Exercise 3

Ask AI to create a faceted line graph showing the rank of the songs "Dancing Queen", "Loser", and "Try Again" from `billboard` over their first 5 weeks, with one panel per track using `facet_wrap()`. Assign the result to `line_graph`. Add this code to a new code chunk in your QMD. In a bash terminal, run `quarto render analysis.qmd`. Check your Live Server tab.

CP/CR.

```{r billboard-100-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

```{r billboard-100-3-test}
line_graph
```

Your code should look something like this:

```
line_graph <- billboard |>
  filter(track %in% c("Dancing Queen", "Loser", "Try Again")) |>
  pivot_longer(
    cols = wk1:wk5,
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE
  ) |>
  ggplot(aes(x = week, y = rank, color = track, group = track)) +
  geom_line() +
  facet_wrap(~ track)
```

For categorical variables, splitting your plot into **facets** creates subplots that each display one subset of the data. The pipe `|>` passes the left side as input to the right side function — use `|>` between data steps and `+` between ggplot layers.

### Exercise 4

Ask AI to create a box plot from `billboard` showing the distribution of ranks of 3 songs. Add this code to a new code chunk in your QMD. In a bash terminal, run `quarto render analysis.qmd`. Check your Live Server tab.

CP/CR.

```{r billboard-100-4}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

```{r billboard-100-4-test}
billboard |>
  filter(track %in% c("Kryptonite", "Loser", "Try Again")) |>
  pivot_longer(
    cols = wk1:wk76,
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE
  ) |>
  ggplot(aes(x = track, y = rank)) +
  geom_boxplot()
```

###

To visualize the relationship between a numerical and a categorical variable, we can use side-by-side box plots. A boxplot shows the distribution's key statistics: the median (middle line), quartiles (box edges), and potential outliers (individual points).

### Exercise 5

Ask AI to create a bar chart from `billboard` showing the count of weeks 3 songs were ranked in the top 100. Add this code to a new code chunk in your QMD. In a bash terminal, run `quarto render analysis.qmd`. Check your Live Server tab.

CP/CR.

```{r billboard-100-5}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

```{r billboard-100-5-test}
billboard |>
  filter(track %in% c("Kryptonite", "Loser", "Try Again")) |>
  pivot_longer(
    cols = wk1:wk76,
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE
  ) |>
  filter(rank <= 100) |>
  ggplot(aes(x = track)) +
  geom_bar()
```

###

**ggplot2** bar charts come in two flavors: one where the bar height is automatically counted from the data, and one where the bar height represents a value already in the data. Make sure to tell AI which you want.

### Exercise 6

Use the following prompt with your AI to build a cleaned version of `billboard`. Add the resulting code to a new code chunk in your QMD. In a bash terminal, run `quarto render analysis.qmd`.

In the R Terminal, run:

```
show_file("analysis.qmd", chunk = "Last")
```

CP/CR.

Here's the prompt we used:

Write R code that starts with `billboard` and pipes it through several operations: filter to keep only rows where `wk1` is less than 101, arrange in ascending order by `wk1`, create a new column called `improvement` equal to `wk1` minus `wk2` placed after the `wk2` column using `mutate()` with `.after = wk2`, select only the columns from `artist` through `improvement`, and rename `wk1` to `week1` and `wk2` to `week2`.

```{r billboard-100-6}
question_text(NULL,
    answer(NULL, correct = TRUE),
    allow_retry = TRUE,
    try_again_button = "Edit Answer",
    incorrect = NULL,
    rows = 6)
```

###

Our result:

```{r billboard-100-6-test}
billboard |>
  filter(wk1 < 101) |>
  arrange(wk1) |>
  mutate(improvement = wk1 - wk2, .after = wk2) |>
  select(artist:improvement) |>
  rename(week1 = wk1, week2 = wk2)
```

If your result doesn't match ours, replace your chunk code with the following before moving on:

<pre><code>```{r}
billboard |>
  filter(wk1 < 101) |>
  arrange(wk1) |>
  mutate(improvement = wk1 - wk2, .after = wk2) |>
  select(artist:improvement) |>
  rename(week1 = wk1, week2 = wk2)
```</code></pre>

`filter()` selects rows by condition, `arrange()` reorders them, `mutate()` creates new columns, `select()` keeps specified columns, and `rename()` changes column names — all without modifying the original tibble.

### Exercise 7

Within the latest code chunk, add the option: `#| cache: true`. Assign the result of the pipe to `billboard_clean`. If you have not done so already, add `analysis_cache` to your `.gitignore`.

In a bash terminal, run `quarto render analysis.qmd`. This first render will take some time — by including `#| cache: true`, Quarto saves the result to disk so the next render loads from cache instead of re-running. Check your Live Server tab.

In the R Terminal, run:

```
show_file("analysis.qmd", chunk = "Last")
```

CP/CR.

```{r billboard-100-7}
question_text(NULL,
    answer(NULL, correct = TRUE),
    allow_retry = TRUE,
    try_again_button = "Edit Answer",
    incorrect = NULL,
    rows = 8)
```

###

The pipe takes the thing on its left and passes it along to the function on its right so that `billboard |> filter(wk1 < 101)` is equivalent to `filter(billboard, wk1 < 101)`. The easiest way to pronounce the pipe is "then".

### Exercise 8

Ask AI to pipe `billboard_clean` into a histogram of `improvement` with a title, subtitle, and caption. Adjust the theme so the title and subtitle are readable. Add this code to a new code chunk in your QMD. In a bash terminal, run `quarto render analysis.qmd`. Check your Live Server tab.

In the R Terminal, run:

```
show_file("analysis.qmd", chunk = "Last")
```

CP/CR.

```{r billboard-100-8}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 12)
```

###

```{r billboard-100-8-test}
ggplot(billboard_clean, aes(x = improvement)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") +
  labs(
	title = "Distribution of Billboard Top 100 Song Ranking Improvements From Week 1 to Week 2",
	subtitle = "Most songs either had no improvement or improved their ranking by up to 20 places",
	x = "Song Rank Improvement",
	y = "Count",
	caption = "Source: Song rankings for Billboard top 100 in the year 2000"
  ) +
  theme(
    plot.title = element_text(size = 7.5, face = "bold"),
    plot.subtitle = element_text(size = 8)
  )
```

###

A histogram divides the x-axis into equally spaced bins and uses the bar height to display observation counts. Many `theme_*()` functions exist within **ggplot2** that create different appearance themes for plots.

## Music
###

The **music** CSV file comes from the [**Million Song Dataset**](https://corgis-edu.github.io/corgis/csv/music/), which used a company called the Echo Nest to derive data points about one million popular contemporary songs. **Music** only contains 10,000 rows of data of the songs from the **Million Song Dataset**.

Key functions you'll learn:

[`read_csv()`](https://readr.tidyverse.org/reference/read_delim.html) - Reads and loads CSV files

[`write_csv()`](https://readr.tidyverse.org/reference/write_delim.html) - Writes a CSV file

[`pivot_longer()`](https://tidyr.tidyverse.org/reference/pivot_longer.html) - Transforms data from a "wide" to a "long" format

[`pivot_wider()`](https://tidyr.tidyverse.org/reference/pivot_wider.html) - Transforms data from a "long" to a "wide" format


### Exercise 1

We begin by downloading our CSV file `music.csv` to the `data` directory directly from the internet using `download.file()`.

In the R Terminal, run:

```
download.file(
  "https://corgis-edu.github.io/corgis/datasets/csv/music/music.csv",
  destfile = "data/music.csv",
  mode = "wb"
)
```

CP/CR.

```{r music-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

"CSV" stands for **C**omma-**S**eparated **V**alues, meaning that the variable names and data values are separated by commas in the file. CSV files are just one type of text file. A text file is any file which includes plain text. The contents of such files are easy to look at in any text editor, or in VSCode.

### Exercise 2

Ask AI to read a `.csv` dataset called `data/music.csv` using `read_csv()`. Add this code to a new code chunk in your QMD. Make sure the result of the function is not assigned to a variable. In a bash terminal, run `quarto render analysis.qmd`. Check your Live Server tab.

CP/CR.

```{r music-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```{r music-2-test}
read_csv("data/music.csv")
```

###

When you run `read_csv()`, it prints information about the rows, columns, and data types it detected. The **readr** package includes similar functions for other file formats like `read_tsv()` for tabs and `read_delim()` for custom separators.

### Exercise 3

If using a chat AI, share the code from above and the `music.csv` tibble from the R Terminal.

Edit the code so that the `read_csv()` skips the first 5 lines. Continue the code so that the `read_csv()` will also not read the first row as headings.

Copy and paste the AI generated code in the place to edit the previous code. In a bash terminal, run `quarto render analysis.qmd`. Check your Live Server tab.

CP/CR.

```{r music-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

```{r music-3-test}
read_csv("data/music.csv", skip = 5, col_names = FALSE)
```

###

The argument `skip` is used to skip rows, but to skip columns, you can use the `col_only()` function as the argument to `col_types` in order to read in only the columns which you want. The argument `col_names` can also be used to create custom column names. There isn't anything wrong with the first five lines of this file.

### Exercise 4

Edit the code so that `read_csv()` is assigned to an object `df` and told that the column type of `artist.familiarity` is a double. Then check `df` for any parsing failures.

If you're using a chat AI, provide the `music.csv` tibble again and the previous code.

Copy and paste the AI generated code in the place to edit the previous code. In a bash terminal, run `quarto render analysis.qmd`. Check your Live Server tab.

CP/CR.

```{r music-4}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

```{r music-4-test}
df <- read_csv(
   "data/music.csv",
   col_types = cols(
     artist.familiarity = col_double()
   )
)

problems(df)
```

###

CSV files don't specify data types, so **readr** automatically guesses whether columns contain text, numbers, or other values. The `col_types` argument lets you manually specify column types when readr guesses incorrectly.

### Exercise 5

Now delete `df` along with the code. The chunk should be empty.

Edit the code `read_csv("data/music.csv")` so that it only reads the columns `artist.familiarity`, `artist.hotttnesss`, and `artist.id`, overriding their types to be strings.

If you're using a chat AI, provide the `music.csv` tibble again and the previous code.

Copy and paste the AI generated code into the empty chunk. In a bash terminal, run `quarto render analysis.qmd`. Check your Live Server tab.

CP/CR.

```{r music-5}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

```{r music-5-test}
read_csv(
  "data/music.csv",
  col_types = cols_only(
    artist.familiarity = col_character(),
	artist.hotttnesss = col_character(),
	artist.id = col_character()
  )
)
```

###

**readr** lets you specify the type of each column — text, numeric, integer, logical — when importing data. This is useful when **readr** guesses incorrectly or when you need to enforce specific types.

### Exercise 6

Assign this code to the object `music_1`. If using a chat AI, share a few rows of `music_1` from the R Terminal.

Based on the `music_1` tibble previously provided, create a similar tibble with only 3 rows of data using `tibble()`, and with the same 3 columns. Don't assign the tibble to an object.

Copy and paste the AI generated code below the previous code. In a bash terminal, run `quarto render analysis.qmd`. Check your Live Server tab.

CP/CR.

```{r music-6}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

```{r music-6-test}
tibble(
  artist.familiarity = c("0.85", "0.67", "0.92"),
  artist.hotttnesss  = c("0.78", "0.55", "0.88"),
  artist.id          = c("AR123", "AR456", "AR789")
)
```

###

Sometimes you'll need to assemble a tibble "by hand," doing a little data entry in your R script. **tibble** provides two ways to do this: laying out the data column by column, or row by row.

### Exercise 7

Continue the previous code with the new tibble so that it is fed through a pipe and is written into a new CSV file titled `music_fake.csv` within the `data` directory.

If you're using a chat AI, provide the previous code.

Manually, using the same command, do the same for the `music_1` tibble.

Copy and paste the AI generated code into the appropriate place. In a bash terminal, run `quarto render analysis.qmd`. Check your Live Server tab.

CP/CR.

```{r music-7}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

```{r music-7-test}
tibble(
  artist.familiarity = c("0.85", "0.67", "0.92"),
  artist.hotttnesss  = c("0.78", "0.55", "0.88"),
  artist.id          = c("AR123", "AR456", "AR789")
) |>
  write_csv("data/music_fake.csv")

write_csv(music_1, "data/music_1.csv")
```

###

**readr** can also write data back to disk as CSV or TSV files. You can control how missing values are written and whether to append to an existing file.

### Exercise 8

Creating a vector of both CSV files, use `read_csv()` to read both CSV files `"music_1.csv"` and `"music_fake.csv"` and create a new column showing the file that the data comes from with the argument `id`. Don't assign the combined data frame to an object.

If you're using a chat AI, provide the `music.csv` tibble again and the previous code.

Copy and paste the AI generated code in the place following the previous code. In a bash terminal, run `quarto render analysis.qmd`. Check your Live Server tab.

CP/CR.

```{r music-8}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

```{r music-8-test}
read_csv(c("data/music_1.csv", "data/music_fake.csv"), id = "file")
```

###

With `read_csv()` you can read multiple CSV files at once and stack them on top of each other in a single data frame. The code above will only work if you have the CSV files in a data folder in your project.

###

Now, delete all code regarding the CSV file.

We will now learn how to make our data tidy.

### Exercise 9

Use the following prompt with your AI to create a tidy version of `billboard`. Add the resulting code to a new code chunk in your QMD. In a bash terminal, run `quarto render analysis.qmd`.

In the R Terminal, run:

```
show_file("analysis.qmd", chunk = "Last")
```

CP/CR.

Here's the prompt we used:

Write R code that starts with `billboard` and pipes it through `pivot_longer()` to collect all columns starting with "wk" into a `week` column and a `rank` column, dropping NA values with `values_drop_na = TRUE`. Continue the pipe with `mutate()` to convert `week` from a character string to a number using `parse_number()`.

```{r music-9}
question_text(NULL,
    answer(NULL, correct = TRUE),
    allow_retry = TRUE,
    try_again_button = "Edit Answer",
    incorrect = NULL,
    rows = 6)
```

###

Our result:

```{r music-9-test}
billboard |>
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE
  ) |>
  mutate(week = parse_number(week))
```

If your result doesn't match ours, replace your chunk code with the following before moving on:

<pre><code>```{r}
billboard |>
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE
  ) |>
  mutate(week = parse_number(week))
```</code></pre>

[`pivot_longer()`](https://tidyr.tidyverse.org/reference/pivot_longer.html) "lengthens" data by turning column names into a variable (`names_to`) and values into another (`values_to`). `parse_number()` extracts the first number from a string, making `week` numeric for plotting.

### Exercise 10

In the R Terminal, type `household` and press `Enter`. The result should be a tibble that is a part of **tidyverse**.

If using a chat AI, share the first 3 lines of the `household` tibble from the R Terminal.

Using the `household` tibble, create code that changes the columns so that there are only 4 columns, `family`, `child`, `dob`, and `name`.

Copy and paste the AI generated code in the place following the previous code. In a bash terminal, run `quarto render analysis.qmd`. Check your Live Server tab.

CP/CR.

```{r music-10}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

```{r music-10-test}
household |>
  pivot_longer(
    cols = !family,
    names_to = c(".value", "child"),
    names_sep = "_child"
  )
```

###

`pivot_longer()` makes data longer by turning column names into variables, with `.value` letting part of a name become a variable.
`pivot_wider()` does the reverse, spreading rows into columns using `values_from`, `names_from`, and `id_cols` to keep rows unique.

### Exercise 11

Within the latest code chunk, add the option: `#| cache: true`. Assign the result of the pipe to `billboard_longer`. If you have not done so already, add `analysis_cache` to your `.gitignore`.

In a bash terminal, run `quarto render analysis.qmd`. This first render will take some time — by including `#| cache: true`, Quarto saves the result to disk so the next render loads from cache instead of re-running. Check your Live Server tab.

In the R Terminal, run:

```
show_file("analysis.qmd", chunk = "Last")
```

CP/CR.

```{r music-11}
question_text(NULL,
    answer(NULL, correct = TRUE),
    allow_retry = TRUE,
    try_again_button = "Edit Answer",
    incorrect = NULL,
    rows = 8)
```

###

There are three interrelated rules that make a dataset tidy: each variable is a column, each observation is a row, and each value is a cell. Most real-world data is untidy because it's often organized for data entry rather than analysis.

### Exercise 12

Generate R code that uses `billboard_longer` to plot a basic graph showing the Billboard rank of each track over the weeks after they were released. If using a chat AI, copy/paste a few rows of `billboard_longer` from the R Terminal to give it context. In-environment agents can access the data directly.

Within `labs()`, edit or add a proper title, subtitle, and caption. Don't assign the code for the plot to any variable. Put the plot code in a new code chunk. In a bash terminal, run `quarto render analysis.qmd`. Check your Live Server tab. Make your plot look nice.

In the R Terminal, run:

```
show_file("analysis.qmd", chunk = "Last")
```

CP/CR.

```{r music-12}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 12)
```

###

```{r music-12-test}
billboard_longer |>
  ggplot(aes(x = week, y = rank, group = track)) +
  geom_line(alpha = 0.2) +
  scale_y_reverse() +
  labs(
    title = "Billboard Track Rankings Over Time",
	subtitle = "Most tracks rise in Billboard rankings in their first 10 weeks",
    x = "Weeks Since Release",
    y = "Billboard Rank",
	caption = "Source: billboard from tidyr"
  )
```

###

Most real analyses will require at least a little tidying. You'll begin by figuring out what the underlying variables and observations are, then pivot your data into a tidy form with variables in the columns and observations in the rows.


## Summary
###

This section covered key concepts from [Chapter 1: Data visualization](https://r4ds.hadley.nz/data-visualize.html), [Chapter 3: Data transformation](https://r4ds.hadley.nz/data-transform.html), [Chapter 5: Data tidying](https://r4ds.hadley.nz/data-tidy.html), and [Chapter 7: Data import](https://r4ds.hadley.nz/data-import.html) from [*R for Data Science (2e)*](https://r4ds.hadley.nz/) by Hadley Wickham, Mine Çetinkaya-Rundel, and Garrett Grolemund.

You will learn about importing, organizing, and visualizing data using core packages such as **[readr](https://readr.tidyverse.org/)**, **[dplyr](https://dplyr.tidyverse.org/)**, and **[ggplot2](https://ggplot2.tidyverse.org/)**. Important functions include: [`read_csv()`](https://readr.tidyverse.org/reference/read_delim.html), [`filter()`](https://dplyr.tidyverse.org/reference/filter.html), [`ggplot()`](https://ggplot2.tidyverse.org/reference/ggplot.html), and [`pivot_longer()`](https://tidyr.tidyverse.org/reference/pivot_longer.html).

### Exercise 1

In a bash terminal, run `quarto render analysis.qmd` to ensure that everything works. Check your Live Server tab — the resulting HTML page should be attractive, showing clean versions of your two plots.

In the R Terminal, run:

```
tutorial.helpers::show_file("analysis.qmd")
```

CP/CR.

```{r summary-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 30)
```

###

Tabular data is considered *tidy* if and only if it satisfies the following rules:

1.  Each variable is a column; each column is a variable.
2.  Each observation is a row; each row is an observation.
3.  Each value is a cell; each cell is a single value.

### Exercise 2

Publish your rendered QMD to GitHub Pages. In a bash terminal, run:

````
quarto publish gh-pages analysis.qmd
````

If the bash terminal is still running from rendering, stop it first using `Ctrl/Cmd + C`.

Copy/paste the resulting URL below.

```{r summary-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 1)
```

###

Publishing the QMD creates a public URL anyone can use to view the rendered project.

### Exercise 3

Commit and push all your files. Copy/paste the URL to your Github repo.

```{r summary-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

GitHub repositories are where people can view the code you used to create your project.

###

`pivot_longer()` and `pivot_wider()` are complicated and unique functions, and you can learn more about how they work step by step with the book we used to create this tutorial. The sections for [`pivot_longer()`](https://r4ds.hadley.nz/data-tidy.html#how-does-pivoting-work) and [`pivot_wider()`](https://r4ds.hadley.nz/data-tidy.html#how-does-pivot_wider-work) will explain in depth.

```{r download-answers, child = system.file("child_documents/download_answers.Rmd", package = "tutorial.helpers")}
```
